---
title: "Exploratory Data Analaysis 2 with NBA Dataset (Using ggplot and dplyr)"
author: "Hubert Luo"
date: "February 22, 2018"
output: github_document
---

```{r}
library(ggplot2)
library(dplyr)
library(readr)
dat <- read_csv('nba2017-players.csv')
```

##Filtering, Slicing, and Selecting (Standard and Piping Syntax)

The following commands contain 3 components:
1. Description of command
2. Command with regular syntax
3. Command with piping syntax

```{r}
#use slice() to subset the data by selecting the first 5 rows.
slice(dat,1:5)
dat %>% slice(1:5)
#use slice() to subset the data by selecting rows 10, 15, 20, ..., 50.
slice(dat,2:10*5)
dat %>% slice(2:10*5)
#use slice() to subset  the data by selecting the last 5 rows.
slice(dat,(nrow(dat)-4):nrow(dat))
dat %>% slice(nrow(dat)-4:nrow(dat))
#use filter() to subset those players with height less than 70 inches tall.
filter(dat, height < 70)
dat %>% filter(height < 70)
#use filter() to subset rows of Toronto Raptors ('TOR').
filter(dat, team == "TOR")
dat %>% filter(team == "TOR")
#use filter() to subset rows of TOR centers ('C').
filter(dat, team == "TOR" & position == "C")
dat %>% filter(team == "TOR" & position == "C")
#use filter() and then select(), to subset rows of lakers ('LAL'), and then display their names.
select(filter(dat, team == "LAL"),player)
dat %>% filter(team == "LAL") %>% select(player)
#use filter() and then select(), to display the name and salary, of TOR point guards
select(filter(dat, team == "TOR" & position == "PG"), player, salary)
dat %>% filter(team == "TOR" & position == "PG") %>%  select(player, salary)
#select the name, age, and team, of players with more than 10 years of experience, making 10 million dollars or less.
select(filter(dat, experience > 10 & salary <= 10e6), player, age, team)
dat %>% filter(experience > 10 & salary <= 10e6) %>% select(player, age, team)
#select the name, team, height, and weight, of rookie players, 20 years old, displaying only the first five occurrences (i.e. rows)
slice(select(filter(dat, experience == 0 & age == 20), player, team, height, weight),1:5)
dat %>% filter(experience == 0 & age == 20) %>% select(player, team, height,weight) %>% slice(1:5)
```

##Add New Variables by mutate and Reordering Rows by arrange

```{r}
#Create data frame of only players from the Toronto Raptors
tor <- filter(dat, team == 'TOR')
tor

#Add column of height/weight
tor2 = mutate(tor, ht_wt = height/weight)
tor2

#Arrange by height, ascending and descending
arrange(tor, height)
arrange(tor,desc(height))

#using the data frame tor, add a new variable product with the product of height and weight and show the 5 players with lowest product
tor %>% mutate(product = height*weight) %>% arrange(product) %>% select(player,position,product) %>% slice(1:5)

#create a new data frame gsw3, by adding columns log_height and log_weight with the log transformations of height and weight.
tor3 = mutate(tor, log_height = log(height), log_weight = log(weight))

#use the original data frame to filter() and arrange() those players with height less than 71 inches tall, in increasing order.
arrange(filter(dat, height < 71),height)
dat %>% filter(height < 71) %>% arrange(height)

#display the name, team, and salary, of the top-5 highest paid players
select(slice(arrange(dat, desc(salary)),1:5),c(player, team, salary))
dat %>% arrange(desc(salary)) %>% slice(1:5) %>% select(player, team, salary)

#display the name, team, and points3, of the top 10 three-point players
select(slice(arrange(dat, desc(points3)),1:10),c(player, team, points3))
dat %>% arrange(desc(points3)) %>% slice(1:10) %>% select(player, team, points3)

#create a data frame gsw_mpg of GSW players, that contains variables for player name, experience, and min_per_game (minutes per game), sorted by min_per_game (in descending order)
gsw_mpg = arrange(select(mutate(filter(dat,team=="GSW"),min_per_game = minutes/games),c(player, experience, min_per_game)),desc(min_per_game))
gsw_mpg
```

###Summarizing values with summarise and Grouped Operations

```{r}
# average salary of NBA players
summarise(dat, avg_salary = mean(salary))

#use summarise() to get the largest height value.
summarise(dat,max_height = max(height))

#use summarise() to get the standard deviation of points3.
summarise(dat,sd_3pt = sd(points3))

#use summarise() and group_by() to display the median of three-points, by team.
summarise(group_by(dat, team),median_3pt = median(points3))
dat %>% group_by(team) %>% summarise(median_3pt = median(points3))

#display the average triple points by team, in ascending order, of the bottom-5 teams (worst 3pointer teams)
slice(arrange(summarise(group_by(dat,team),average_3pt = mean(points3)),average_3pt),1:5)
dat %>% group_by(team) %>% summarise(average_3pt = mean(points3)) %>% arrange(average_3pt) %>% slice(1:5)

#obtain the mean and standard deviation of age, for Power Forwards, with 5 and 10 years (including) years of experience.
summarise(filter(dat, position =="PF" & experience>=5 & experience <= 10),average_age = mean(age), sd_age = sd(age))
dat %>% filter(position == "PF" & experience >= 5 & experience <= 10) %>% summarise(average_age = mean(age), sd_age = sd(age))
```

###More Dataframe Manipulation

```{r}
#Display those rows associated to players having height less than 70 inches tall.
show(dat[dat$height<70,])
dat %>% filter(height < 70)

#Display the names and salaries of centres with over 10 years of playing experience
show(dat[dat$position=='C' & dat$experience > 10,c('player','salary')])
dat %>% filter(position == 'C' & experience > 10) %>% select(player, salary)

#Create a data frame UCLA with the data of players from college UCLA ("University of California, Los Angeles").
ucla <- data.frame(dat[dat$college == "University of California, Los Angeles",])
ucla <- dat%>%filter(college == "University of California, Los Angeles")

#Create a data frame rookies with those players with 0 years of experience.
rookies <- data.frame(dat[dat$experience == 0,])
rookies <- dat%>%filter(experience==0)

#Create a data frame rookie_centers with the data of Center rookie players.
rookie_centres <- data.frame(rookies[rookies$position == 'C',])
rookie_centres <- rookies%>%filter(position=='C')

#Create a data frame top_players for players with more than 50 games and more than 300 minutes played.
top_players <- dat[(dat$games > 50) & (dat$minutes > 300),]
top_players <- dat%>%filter(games>50 & minutes > 300)

#What's the largest height value?
max(dat$height)
dat %>% summarise(max(height))

#What's the minimum height value?
min(dat$height)
dat %>% summarise(min(height))

#What's the overall average height?
mean(dat$height)
dat %>% summarise(mean(height))

#Who is the tallest player?
dat[dat$height == max(dat$height),'player']
dat %>% arrange(desc(height)) %>% select(player) %>% slice(1)

#Who is the shortest player?
dat[dat$height == min(dat$height),'player']
dat %>% arrange(height) %>% select(player) %>% slice(1)

#Which are the unique teams?
unique(dat$team)
dat %>% distinct(team)

#How many different teams?
length(unique(dat$team))
dat %>% distinct(team) %>% nrow

#Who is the oldest player?
dat[dat$age == max(dat$age),'player']
dat %>% arrange(desc(age)) %>%  select(player) %>% slice(1)

#What is the median salary of all players?
median(dat$salary)
dat %>% summarise(median(salary))

#What is the median salary of the players with 10 years of experience or more?
median(dat[dat$experience >= 10, 'salary'][[1]])
dat %>% filter(experience >= 10) %>% summarise(median(salary))
  
#What is the median salary of Shooting Guards (SG) and Point Guards (PG)?
median(dat[(dat$position == 'PG') | (dat$position == 'SG'),'salary'][[1]])
dat %>% filter(position == 'PG' | position == 'SG') %>% summarise(median(salary))

#What is the median salary of Power Forwards (PF), 29 years or older, and 74 inches tall or less?
median(dat[(dat$position == 'PF') & (dat$age >= 29) & (dat$height <= 74),'salary'][[1]])
dat %>% filter(position == 'PG' & age >= 29 & height <= 74) %>% summarise(median(salary))

#How many players are from "University of California, Berkeley"?
nrow(na.omit(dat[dat$college == "University of California, Berkeley",]))
dat %>% filter(college == "University of California, Berkeley") %>% nrow

#Are there any players with weight greater than 260 pounds? If so how many and who are they?
nrow(dat[dat$weight > 260,])
dat[dat$weight > 260,'player']
dat %>% filter(weight > 260) %>% select('player')

#How many players did not attend a college in the US?
nrow(dat[is.na(dat$college),])
dat %>% filter(is.na(college)) %>% nrow

#Who is the player with the maximum rate of points per minute?
dat[dat$points/dat$minutes == max(dat$points/dat$minutes),'player']
dat %>% arrange(desc(points/minutes)) %>% select(player) %>% slice(1)

#Who is the player with the maximum rate of 3 pointers per minute?
dat[dat$points3/dat$minutes == max(dat$points3/dat$minutes),'player']
dat %>% arrange(desc(points3/minutes)) %>% select(player) %>% slice(1)
```

###Grouping the Data

```{r}
#Create a data frame with the average height, average weight, and average age, grouped by position
aggregate(dat[,c("height","weight","age")],by = list(dat$age),FUN = mean)
dat %>% group_by(position) %>% summarise(mean(height),mean(weight),mean(age))

#Create a data frame with the average height, average weight, and average age, grouped by team
aggregate(dat[,c("height","weight","age")],by = list(dat$team),FUN = mean)
dat %>% group_by(team) %>% summarise(mean(height),mean(weight),mean(age))

#Create a data frame with the average height, average weight, and average age, grouped by team and position.
aggregate(dat[,c("height","weight","age")],by = list(dat$team, dat$position),FUN = mean)
dat %>% group_by(position, team) %>% summarise(mean(height),mean(weight),mean(age))

#Create a data frame with the minimum salary, median salary, mean salary, and maximum salary, grouped by team and position.
nba_summary <- data.frame(aggregate(dat[,c("salary")],by = list(dat$team, dat$position),FUN = min), 
           aggregate(dat[,c("salary")],by = list(dat$team, dat$position),FUN = median),
           aggregate(dat[,c("salary")],by = list(dat$team, dat$position),FUN = mean),
           aggregate(dat[,c("salary")],by = list(dat$team, dat$position),FUN = max))[,c(1,2,3,6,9,12)]
colnames(nba_summary) <- c("Team","Position","Min Salary", "Median Salary", "Mean Salary", "Max Salary")
nba_summary
dat %>% group_by(position, team) %>% summarise(min(salary),median(salary),mean(salary),max(salary))
```

###Visualizations with ggplot

```{r scatterplot}
# scatterplot (option 1)
ggplot(data = dat) +
  geom_point(aes(x = points, y = salary))
```

```{r scatterplot2}
# scatterplot (option 2)
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point()
```

```{r coloured_scatterplot}
# colored scatterplot 
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position))
```

```{r size_colour_scatterplot}
# sized and colored scatterplot 
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position, size = points3))
```

```{r transparent_scatterplot}
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position, size = points3), alpha = 0.7)
```

```{r gsw_scatterplot}
#Use the data frame tor to make a scatterplot of height and weight.
ggplot(data = tor, aes(x = height, y = weight)) +
  geom_point()
```